\section{Developer Install%
\label{sec:dev-install}}

\subsection{Installation and setup}

To install the DPJ compiler and runtime from source, do the following.
%
\begin{enumerate}
%
\item Check that your system meets the requirements in
  \S~\ref{sec:requirements}.
%
\item Get the source code distribution.  We recommend using Eclipse to
  check out the source code; that way you'll have access to all of
  Eclipse's code browsing and editing features.  Go to
  \kwd{http://www.eclipse.org} to get Eclipse.  Then you'll need to install
  EGit (Git for Eclipse).  This tutorial tells you how to do it:
%
\begin{description}
\item \kwd{http://www.vogella.de/articles/EGit/article.html}
\end{description}
%
Then you can check out the DPJ source tree from inside Eclipse, again
as explained in the tutorial.  Use this as the repository URL:
%
\begin{description}
\item \kwd{git://github.com/dpj/DPJ.git}
\end{description}
%
The following steps assume you are calling the root of the working
directory \kwd{DPJ} and putting it in your home directory.  If not,
make the appropriate adjustments.

If you don't want to use Eclipse, you can check out the DPJ source
tree from the command line:
%
\begin{verbatim}
cd ~
git clone git://github.com/dpj/DPJ.git DPJ
\end{verbatim}
%
Again, make the appropriate adjustments if you are putting the root of
the working directory somewhere else.
%
\item Set the \kwd{DPJ\_ROOT} environment variable.  For example, if
  the root of your DPJ tree is \kwd{DPJ} in your home directory and
  you are using the C shell, put this in your \kwd{.cshrc} file:
%
\begin{verbatim}
setenv DPJ_ROOT ~/DPJ
\end{verbatim}
%
Then do
%
\begin{verbatim}
source ~/.cshrc
\end{verbatim}
%
\item Tell \kwd{ant} to build the compiler, using your JDK: 
%
\begin{verbatim}
cd ~/DPJ/Implementation/Compiler/make
ant -Dboot.java.home=/usr/lib/jvm/java
\end{verbatim}
%
If your JDK is not located at \kwd{/usr/bin/jvm/java}, then substitute
the appropriate path.
%
\item A successful compiler build puts \kwd{jar} files for the
  compiler and related tools (such as \kwd{javadoc}) in
\begin{description}
\item \kwd{Implementation/Compiler/build}.  
\end{description}
The executable files \kwd{dpjc} and \kwd{dpj} in
\kwd{Implementation/bin} invoke these \kwd{jar} files.
%
\item Add the DPJ compiler bin to your \kwd{PATH}:
\begin{verbatim}
setenv PATH ${PATH}:${DPJ_ROOT}/Implementation/bin/
\end{verbatim}
\item Build the runtime classes located at
  \kwd{{\$\{DPJ\_ROOT\}}/Implementation/Runtime}.  You can expect some
  warnings.
\begin{verbatim}
cd ../../Runtime/
make
\end{verbatim}

\item You can now build and run the kernels:
\begin{verbatim}
cd ${DPJ_ROOT}/Benchmarks/Kernels
make
\end{verbatim}

\item Then run all the kernel tests: 
\begin{verbatim}
make test-all
\end{verbatim}


\end{enumerate}

\subsection{Testing the Installation%
\label{sec:testing}}
\begin{enumerate}
\item JUnit 4 tests are in
  \kwd{\$\{DPJ\_ROOT\}/Implementation/Compiler/test/dpj-junit-tests}.  The
  corresponding DPJ source files are in \kwd{test/dpj-programs}.
\item The JUnit tests are designed to be run from within Eclipse, by
  running the launcher file
%
\begin{description}
\item \kwd{test/dpj-junit-tests/dpj-junit-tests.launch}.
\end{description}
%
If the JUnit tests fail with a \kwd{MethodNotFound} error, it is
probably because the classpath is wrong.  Many of the javac classes
are included in the system library (\kwd{rt.jar}), and so the DPJ
versions must precede the system library in the classpath.  If you use
the launcher, this error should not happen.
\end{enumerate}

\subsection{Browsing and Modifying the Compiler Code}

After an initial \kwd{ant} build, DPJ javac will build automatically
in Eclipse (albeit to a \kwd{bin/} directory instead of \kwd{build/}).
However, if any of the resource bundles are changed (be sure to change
the \kwd{.properties} files in
%
\begin{description}
\item \kwd{src/share/classes/com/sun/tools/javac/resources} 
\end{description}
and not the generated \kwd{.java} files!), the compiler must be
re-built using the \kwd{ant} script.
 
The \kwd{ant} script bootstraps the compiler.  So if there is an
internal compiler error, it is due to a fault in the DPJ compiler.
Note also that the compiler may compile successfully in Eclipse but
fail when it is bootstrapped in the \kwd{ant} script due to differences
between pure Java and DPJ (or an error in the DPJ compiler).  In
particular, note that the following keywords are reserved in DPJ and
therefore may not be used as variable names (in any DPJ program,
including the DPJ compiler):
\begin{itemize}
     \item \kwd{under}
     \item \kwd{region}
     \item \kwd{reads}
     \item \kwd{writes}
\end{itemize}



