\section{The DPJ Runtime%
\label{sec:runtime}}
\cutname{runtime.html}

This section gives an overview of the classes \kwd{DPJArray} and
\kwd{DPJPartition} in the package \kwd{DPJRuntime}.  These classes are
part of the runtime supplied with DPJ and are useful for manipulating
arrays.  For full documentation of these classes, and for coverage of
the other DPJ runtime classes, see the HTML documentation included in
the DPJ release at 
%
\begin{description}
\item \kwd{DPJ/Implementation/Runtime/dpjdoc}.
\end{description}

To use the DPJ runtime, you have to do the following:
%
\begin{enumerate}
%
\item Put \kwd{import DPJRuntime.*} at the top of your code (or just
  import the class or classes you want to use), as for regular Java.
%
\item When compiling your code, you must compile the runtime classes
  located at \kwd{Implementation/Runtime/dpj} in the DPJ release
  together with any code that uses the runtime.  For example, if class
  file \kwd{Foo.java} depends on the DPJ runtime, then you need to do
  something like this:
%
\begin{description}
\item \kwd{dpj Foo.java \$\{DPJ\_ROOT\}/Implementation/Runtime/dpj}
\end{description}
%
\emph{It is not sufficient to put the runtime classes in your class
  path.}  If you do that, you will get a lot of errors.  The reason is
that DPJ bytecode doesn't yet properly support separate compilation of
DPJ annotations; the DPJ compiler needs the DPJ source for all
annotated code to process the annotations.
%
\item When running your code, put the runtime classes located at
  \kwd{Implementation/Runtime/classes} in your class path, as for
  regular Java.
%
\end{enumerate}

\subsection{\kwd{DPJArray}%
\label{sec:runtime:array}}
\cutname{runtime.array.html}

In parallel algorithms that operate on arrays, especially
divide-and-conquer algorithms, it is often necessary to split an array
into parts and operate separately on the separate parts.  Because this
operation is so useful, some languages (e.g., Fortran 90) include
first-class support for extracting such ``slices'' from arrays.
Unfortunately, Java does not.  Instead of adding slices directly to
arrays, which would require redefining Java array types and array
operations (and would be hard to make consistent with legacy Java),
DPJ provides array slices as an operation of a library class, called
\kwd{DPJArray}.  There are two kinds of \kwd{DPJArray} classes: the
generic \kwd{DPJArray} class, and a set of \kwd{DPJArray} classes
specialized to primitive types.

\bfhead{Generic \kwd{DPJArray}:} The generic \kwd{DPJArray} class
represents an array of objects.  It has one type parameter and one RPL
parameter:
%
\begin{dpjlisting}
class DPJArray<type T, region R>
\end{dpjlisting}
%
The type parameter is the element type of the array, and the RPL
parameter is the RPL of the array storage.

\ithead{Creating a \kwd{DPJArray}:} There are three ways to create a
\kwd{DPJArray} object.  The first is to call a constructor that takes
just a length argument.  This operation creates a new \kwd{DPJArray}
with the specified length, type, and RPL.  For example, the following
code creates a \kwd{DPJArray} storing 10 \kwd{Integer} objects, and
the storage is in RPL \kwd{r}:
%
\begin{dpjlisting}
region r;
DPJArray<Integer,r> A = new DPJArray<Integer,r>(10);
\end{dpjlisting}
%

Creating a fresh DPJ array is useful, but sometimes you have a Java
array and you want to make a \kwd{DPJArray} out of it.  So the second
way to create a \kwd{DPJArray} is to call a constructor that takes an
ordinary Java array as an argument.  For example, the following code
creates a Java array of 10 \kwd{Integer} objects, and wraps it in a
\kwd{DPJArray}:
%
\begin{dpjlisting}
region r;
Integer[]<r> a = new Integer[10]<r>;
DPJArray<Integer,r> A = new DPJArray<Integer,r>(a);
\end{dpjlisting}
%
Note that the type and RPL of the array being passed in (here,
\kwd{Integer} and \kwd{r}) must match the type and RPL argument of the
\kwd{DPJArray} type; if not, a compile error occurs.

The third way to create a \kwd{DPJArray} is to make a \emph{subarray}
(i.e., a slice) of an existing \kwd{DPJArray}.  This is explained
below.

\ithead{Accessing elements:} \kwd{DPJArray} has \kwd{put} and
\kwd{get} operations similar to the ones in \kwd{java.util.ArrayList}.
For example, if \kwd{A1} and \kwd{A2} are \kwd{DPJArray}s, then the
following code fragment gets element 0 out of \kwd{A1} and stores it
into element 0 of \kwd{A2}:
%
\begin{dpjlisting}
A2.put(0, A1.get(0));
\end{dpjlisting}
%
If the \kwd{DPJArray} was created by wrapping a Java array, then the
\kwd{put} operation modifies the wrapped array.  The effect of a
\kwd{get} operation is a read on the region of the \kwd{DPJArray}, and
the effect of a \kwd{put} operation is a write to the region of the
\kwd{DPJArray}.

Accesses are bounds-checked.  Any attempt to access a position less
than 0 or greater than the array length minus one throws an
\kwd{ArrayIndexOutOfBoundsException}.

\ithead{Subarrays:} The real usefulness of \kwd{DPJArray} is its
support for \emph{subarrays}, which are contiguous subsections of an
array.  To create a subarray of a \kwd{DPJArray}, you call the
\kwd{subarray} instance method with a start and length argument.  For
example, the following code creates a \kwd{DPJArray} and then extracts
the subarray of length 2 starting at position 5 (here we use the
default RPL of \kwd{Root}):
%
\begin{dpjlisting}
DPJArray<Integer> A = new DPJArray<Integer>(10);
DPJArray<Integer> B = A.subarray(5,2);
\end{dpjlisting}

There are two nice things about subarrays.  First, creating a subarray
takes minimal time and space overhead.  Nothing is copied, and no new
storage allocated to hold any array elements.  A small object is
created that stores the start position, length, and a reference to the
underlying array.

Second, as far as its API is concerned, a subarray is
indistinguishable from a freshly created \kwd{DPJArray}.  For example,
the subarray created above is zero-indexed, it has length 2, and
attempts to access indices other than 0 and 1 throw an exception.
However, the subarray also provides a view into the original array.
For example, the following code stores 1 into position 0 of \kwd{B},
which is the same as position 5 of \kwd{A}:
%
\begin{dpjlisting}
B.put(0,1);
\end{dpjlisting}
%
You can get the start position out of a \kwd{DPJArray} by reading the
field \kwd{start}, and you can get the length by reading the field
\kwd{length}.

These features allow methods that operate on array subranges to be
zero-indexed, without worrying about the index parameters of the
subranges.  For example, here is some simple recursive code that
increments every position of a \kwd{DPJArray} by 1:
%
\begin{dpjlisting}
<type T, region R>void increment(DPJArray<T,R> A) {
    if (A.length == 0) return;
    if (A.length == 1) {
        A.put(0, A.get(0)+1);
    }
    int mid = A.length / 2;
    increment(A.subarray(0, mid));
    increment(A.subarray(mid, length-mid));
}
\end{dpjlisting}
%
Without the subarray feature, this code would have to be written by
passing the index ranges as parameters to the \kwd{increment} method,
which is ugly.

\bfhead{Specialized \kwd{DPJArray}s:} Because Java does not support
binding primitive types to generic parameters, the DPJ runtime also
has versions of \kwd{DPJArray} specialized to the primitive types
(\kwd{DPJInt} for \kwd{int}, \kwd{DPJBoolean} for \kwd{boolean},
etc.).  These operate identically to the generic \kwd{DPJArray},
except that the element type is given by the class, and there is no
generic parameter.

Another way to write such arrays is to create a generic \kwd{DPJArray}
using the class corresponding to the primitive type.  For example,
instead of a \kwd{DPJArrayInt}, one could use a
\kwd{DPJArray<Integer>}, as described above.  This works, but the code
is more verbose, as well as more memory-intensive and slower, as Java
has to box and unbox all those primitive types when putting them into
and getting them out of the array.


\subsection{\kwd{DPJPartition}%
\label{sec:runtime:partition}}
\cutname{runtime.partition.html}

For parallel divide-and-conquer algorithms on arrays, it is often
important to create disjoint collections of subarrays.  For example, a
parallel sorting algorithm might repeatedly divide an array into
disjoint halves (or quarters, etc.) and operate recursively in
parallel on the pieces.  To support effect checking for this kind of
algorithm, the DPJ runtime includes a class \kwd{DPJPartition} for
representing disjoint collections of subarrays of the same array.
Each of the subarrays in the collection is called a \emph{segment} of
the partition.  As with \kwd{DPJArray}, there is a generic version,
and there are specialized versions.

\bfhead{Generic \kwd{DPJPartition}:}  The generic \kwd{DPJPartition}
class represents an array of subarrays of a \kwd{DPJArray}.  It has
one type parameter and one RPL parameter:
%
\begin{dpjlisting}
class DPJPartition<type T, region R>
\end{dpjlisting}
%
The type parameter is the element type of the \kwd{DPJArray} being
partitioned, and the RPL parameter is the RPL of the array storage.

\ithead{Creating a \kwd{DPJPartition}:} There are several ways to
create a \kwd{DPJPartition}; for a full list, see the HTML
documentation.  Here are two useful ways.  First, \kwd{DPJPartition}
has a constructor that takes a \kwd{DPJArray} to partition, an index
at which to partition, and a \kwd{boolean} value that says whether to
exclude or include the element at the index position.  For example, if
\kwd{A} is a \kwd{DPJArray<Integer>} of length 10, then
%
\begin{dpjlisting}
new DPJArray<Integer>(A, 5, true)
\end{dpjlisting}
%
partitions \kwd{A} into the segments $[0,4]$ and $[6,9]$ (excluding
position 5), while
%
\begin{dpjlisting}
new DPJArray<Integer>(A, 5, false)
\end{dpjlisting}
%
partitions \kwd{A} into the segments $[0,4]$ and $[5,9]$ (including
position 5).  This constructor is useful for parallel
divide-and-conquer algorithms with a fanout of 2.

Second, \kwd{DPJPartition} has a static factory method
\kwd{stridedPartition} that takes a \kwd{DPJArray} to partition and a
stride at which to partition.  For example, if \kwd{A} is a
\kwd{DPJArray<Integer>} of length 10, then
%
\begin{dpjlisting}
new DPJPartition<Integer>(A, 2)
\end{dpjlisting}
%
creates a \kwd{DPJPartition<Integer>} with five segments, each of
length 2.  This feature is useful for parallel divide-and-conquer
algorithms with a fanout of greater than two, as well as flat
partitions (such as tiling an array).

\ithead{Accessing segments:} The field \kwd{length} stores the number
of segments in the partition.  It is \kwd{final}, so reading it has no
effect (\S~\ref{sec:effects:stmt-exp}).  The method \kwd{get} takes an
integer index \kwd{idx} and returns the segment corresponding to that
index (and throws an exception if the index is out of range).  The
type of the segment is \kwd{DPJArray<T,this:[idx]:*>}.

The index-parameterized type returned by \kwd{get} allows the
different segments to be operated on in parallel without interference,
similarly to an index-parameterized array
(\S~\ref{sec:types:array:ipa}).  For example, the following code uses
\kwd{DPJPartition} to parallelize the simple recursive \kwd{increment}
shown in \S~\ref{sec:runtime:array}:
%
\begin{dpjlisting}
<type T, region R>void parallelIncrement(DPJArray<T,R> A) 
  writes R:* {
    if (A.length == 0) return;
    if (A.length == 1) {
        // Effect is 'writes R'
        A.put(0, A.get(0)+1);
    }
    int mid = A.length / 2;
    final DPJPartition<T,R> segs = new DPJPartition<T,R>(A,mid)
    cobegin {
        // Effect is 'writes segs:[0]:*'
        parallelIncrement(segs.get(0));
        // Effect is 'writes segs:[1]:*'
        parallelIncrement(segs.get(1));
    }
}
\end{dpjlisting}
%
For more examples of how to use \kwd{DPJPartition}, see \tutorial.

The variable \kwd{this} in the type gets substituted by the selector
expression at the call site (\S~\ref{sec:types:exp:invoke}), so it is
usually most useful to call \kwd{get} through a \kwd{final} local
variable.  Including the variable in the RPL ensures that the compiler
doesn't erroneously infer disjointness for two different partitions of
the same array.  For example, in the following code,
\kwd{segs1.get(1)} and \kwd{segs2.get(0)} are not disjoint (they
overlap at $[2,7]$):
%
\begin{dpjlisting}
DPJArray<Integer> A = new DPJArray<Integer>(10);
// Create segments [0,1] and [2,9]
DPJPartition<Integer> segs1 = new DPJPartition(A, 2);
// Create segments [0,7] and [8,9]
DPJPartition<Integer> segs2 = new DPJPartition(A, 8);
\end{dpjlisting}

\bfhead{Specialized \kwd{DPJPartition}s:} As with \kwd{DPJArray},
there are versions of \kwd{DPJPartition} specialized to the various
primitive types.  They are provided for convenience and efficiency, as
Java does not support binding primitive types to generic type
parameters.


